Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-11 paper/overview.tex
fixed a TODO

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot, and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s}, and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
nearly linear-time improvement in $m$.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.

@@

Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
computations, gathering them into \texttt{qs2}. This pass is given the boundary
data structure and the rank \texttt{k}, and finishes the computation. Now that
\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
choosing which recursive call of \texttt{qss} is actually relevant for this
\texttt{k}. Since the boundary data structure contains the first-stage data for
\emph{all} of the recursive calls, performing these comparisons essentially
walks the tree, using the rank along with the size data \texttt{i} in order to
look up the \texttt{k}th leftmost node in the tree.

>>>>

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

@@

A first cut at a solution might be to, in a first pass, sort \texttt{l} into
\texttt{s} and then in a second pass, perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $O(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l} and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
nearly linear-time improvement in $m$.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-split}.

@@

Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
computations, gathering them into \texttt{qs2}. This pass is given the boundary
data structure and the rank \texttt{k}, and it finishes the computation. Now that
\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
choosing which recursive call of \texttt{qss} is actually relevant for this
\texttt{k}. Since the boundary data structure contains the first-stage data for
\emph{all} of the recursive calls, performing these comparisons essentially
walks the tree, using the rank along with the size data \texttt{i} in order to
look up the \texttt{k}th leftmost node in the tree.

### 2015-07-11 paper/semantics.tex
fixed a TODO

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with a eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The multistage function \texttt{qss} from \ref{sec:overview} has a
type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
\TODO what?
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val y = ...`
2`    val qs = y`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a function containing no first-stage redexes.

>>>>

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

Multistage functions, such as \texttt{qss} from \ref{sec:overview},
can be represented as terms with a type fitting the pattern \mbox{$A\to \fut (B \to C)$} at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.} 
To apply such a function $f$ to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let fun qs k = 
      prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = 
	  let val y = ... in y
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.
